Making multiple-locus association mapping on a genome-wide scale routine

### 2016-09-20 AM+draft.tex
21/09/16  about to replace introduction with ipad intro

Gwas for many are now the design of choice when interested in elucidating the genetic 
underpinnings  of complex traits.  Over the past decade, gwas have changed 
considerably in focus and scope. Early gwas 
were performed predominantly as case control studies. Their focus 
was on identifying the casual variants underlying common disorders in humans. 
Despite many promising findings, reproducibility of these findings 
proved an issue (refs). With the collection of dense whole genome data challenging, 
early gwas lacked power (refs).  Today, gwas are performed not just in humans but 
animals and plants that may have highly structured pedigrees. 
They can involve hundreds to thousands of individuals with genotypes collected 
from thousands to hundreds of thousands of snps. Where early gwas were interested
 in disease traits, modern gwas are also  interested in  categorical and quantitative traits.

Over the past decade, there have also been impressive advancements in genome wide 
association mapping methods, the statistical techniques for measuring the strength of 
association between a marker locus and trait. Traditionally, association data were 
analysed with simple contingency tests or Cochran-armitage  trend tests. Today, far 
more statistically sophisticated methods are available. One such method is the  QK 
approach, a linear mixed model framework for genome wide association 
mapping (refs).  Several highly efficient computer programs for performing QK-based 
association mapping are available. Programs such as EMMA, fastLMM, GEMMA, and 
XXX(ref) make feasible the analysis of data from hundreds of thousands of snps, collected 
on thousands of subjects. However, not all areas of genome-wide association 
mapping have advanced. It still remains standard practice to analyse genome-wide 
association data with a single locus model. That is,  it is common for the strength of 
association between a marker locus and trait is measured for each locus separately.

>>>>

Over the past decade, g-w association mapping methods have changed considerably. Early mapping methods relied upon 
simple contingency tables or logistic regression for case-control designs and score tests for family-based designs.

Over the past decade, genome-wide association mapping methods, have changed considerably. Early association mapping 
methods were no more complicated than contingency table tests or logistic regression. This is because early association 
studies followed case-control designs. A significant difference in the allele frequency at a locus, between the cases (affecteds) and 
controls (unaffecteds), was interpreted as the locus being in linkage disequilibrium with a casual gene. 
It became known though that differences in allele frequency could also arise due to unrecognised population stratification. 
This prompted a shift towards family-based designs. Family-based designs do not suffer from 
the confounding affects of population stratification. A myriad of new family-based association mapping methods were developed, 
the most popular based on score tests (ref for TDT S-TDT Spielman and Ewens 1998, AJHG 62:450-458, FBAT Naird, NM; Horvath, S; Xu, X 2000, genetic epi V19, Suppl 1 S36-S42).

Over the past decade, genome-wide association mapping, the statistical techniques for analysing data from association studies, has advanced considerably. Early genome-wide association studies (GWAS) began as case-control designs. The premise was that casual genes could be detected via differences in allele frequency between cases (affecteds) and controls (unaffecteds). Association mapping methods for these designs were no more sophisticated than contingency table tests or logistic regression. Unfortunately, differences 
in allele frequency can also arise due to unrecognised population stratification. Consequently, there was a shift towards family-based 
designs. Findings from these designs are
not susceptible to the confounding affects of population stratification. With interest growing in genome-wide association studies, 
a myriad of different association mapping methods for family-based designs were developed.

Over the past decade, there have also been impressive advancements in genome wide 
association mapping methods, the statistical techniques for measuring the strength of 
association between a marker locus and trait. (Needs more detail - what are some of the early assoc study methods (give a single sentence review).... TDT, genomic control, structured association, score tests???? 
Traditionally, association data were 
analysed with simple contingency tests or Cochran-armitage  trend tests. Today, far 
more statistically sophisticated methods are available. One such method is the  QK 
approach, a linear mixed model framework for genome wide association 
mapping (refs).  The QK approach

Several highly efficient computer programs for performing QK-based 
association mapping are available. Programs such as EMMA, fastLMM, GEMMA, and 
XXX(ref) make feasible the analysis of data from large GWAS. 
 However, not all areas of genome-wide association 
mapping have advanced. It still remains standard practice to analyse genome-wide 
association data with a single locus model. That is,  it is common for the strength of 
association between a marker locus and trait is measured for each locus separately.
